---
title: "TRI Data Analysis"
output: html_document
date: "Dic 1"
author: Isabella Gonzalez, William Deffaa, Rebecca Baldwin, Mehedi Marof 
---

# **Introduction:**

Our team wants to answer the next questions using the Toxics Release Inventory (TRI) of the entire U.S. in 2024.

1.  What industry sector and what companies specifically have released the most, and the least amount of toxins?

2.  What specific chemicals are released the most, and the least? What percentage of these chemicals are United States Clean Air Act Chemicals, and what percentage are not?

3.  What percentage of the chemicals released in the country are transferred off-site? In which neighborhoods are the facilities located that receive these chemical transfers or disposals? Why they are released here?

4.  Do higher toxic releases correlate with higher asthma or cancer rates near these sites?

5.  Which Maryland communities are exposed to the highest levels of toxic pollution, and are Black, low-income, or immigrant neighborhoods disproportionately affected?

## **Libraries:**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)

install.packages("USAboundaries")

library(tidyverse)
library(lubridate)
library(janitor)
library(stringr)
library(tigris)
library(ggplot2)
library(sf)


```

## Data and cleaned data:

```{r}
tri_raw <- read_csv("data/tri.csv", show_col_types = FALSE)

tri_clean <- tri_raw |>
  rename_with(~ iconv(.x, to = "ASCII", sub = "")) |>
  rename_with(~ gsub("^[0-9]+\\.\\s*", "", .x)) |>
  clean_names()

numeric_cols <- tri_clean |>
  select(
    contains("release"),
    contains("transfer"),
    contains("waste"),
    contains("total"),
    matches("^x5"),   
    matches("^x6"),   
    matches("^x7"),   
    matches("^x8"),
    -latitude,          
    -longitude
    ) |>
  names()

tri_clean <- tri_clean |>
  mutate(
    across(
      all_of(numeric_cols),
      ~ suppressWarnings(parse_number(as.character(.x)))
    )
  )

str(tri_clean)
```

## **Basic Data Analysis:**

The TRI dataset contains 77,295 rows and 122 columns, and represents facility-level reports. It includes detailed information on each facility’s location, industrial sector, chemical names, release amounts, transfer activities, and indicators like if a chemical is listed under the U.S. Clean Air Act.

Several columns contain missing data. TRI data are self-reported estimates rather than measured environmental concentrations, and the inventory only includes facilities that exceed reporting thresholds, so it doesn't capture all sources of pollution. The dataset also doesn't contain health outcomes or demographic variables, so to answer questions 3, 4 and 5, we have to combine the TRI dataset with other databases.

The dataset would work to analyzing which chemicals, companies, and industry sectors release the most pollution, and to quantify the share of pollution transferred off-site. However, questions concerning asthma or cancer outcomes, or the racial and socioeconomic composition of affected communities, cannot be answered with TRI data alone and require merging with external health and demographic datasets, like the Census ACS data.

The TRI dataset for this project is limited to the year 2024, meaning it cannot show trends over time or historical pollution patterns. The data only include facilities that meet EPA reporting thresholds and only include listed TRI chemicals.

### QUESTION NO. 1: What industry sector and what companies specifically have released the most, and the least amount of toxins?

#### a) Industry sector

```{r}


industry_totals <- tri_clean |>
  group_by(industry_sector) |>
  summarise(total_industry_release = sum(total_releases, na.rm = TRUE)) |>
  arrange(desc(total_industry_release))

head(industry_totals, 10)

tail(industry_totals, 10)
 
```

#### b) Companies

```{r}
company_totals <- tri_clean |>
  group_by(facility_name) |>
  summarise(total_company_release = sum(total_releases, na.rm = TRUE)) |>
  arrange(desc(total_company_release))

head(company_totals, 10)

tail(company_totals, 10)

```

**Findings:**

### QUESTION NO. 2: What specific chemicals are released the most, and the least? What percentage of these chemicals are United States Clean Air Act Chemicals, and what percentage are not?

#### a) Most released chemicals

```{r}
chemical_totals <- tri_clean |>
  group_by(chemical) |>
  summarise(total_chemical_release = sum(total_releases, na.rm = TRUE)) |>
  arrange(desc(total_chemical_release))

head(chemical_totals, 10)

tail(chemical_totals, 10)

```

**GRAPH**

```{r}
chemical_totals |>
  slice_max(total_chemical_release, n = 10) |>
  ggplot(aes(x = reorder(chemical, total_chemical_release),
             y = total_chemical_release)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Chemicals Released (TRI 2024)",
       x = "Chemical",
       y = "Total Releases")
```

#### b) Percentage Clean Air Act Chemicals

```{r}
top10_chems <- chemical_totals |>
  slice_max(total_chemical_release, n = 10)

top10_with_caa <- top10_chems |>
  left_join(
    tri_clean |> distinct(chemical, clean_air_act_chemical),
    by = "chemical"
  )

num_caa_top10 <- sum(top10_with_caa$clean_air_act_chemical == 1, na.rm = TRUE)
num_non_caa_top10 <- 10 - num_caa_top10

percent_caa_top10 <- (num_caa_top10 / 10) * 100
percent_non_caa_top10 <- 100 - percent_caa_top10

percent_caa_top10
percent_non_caa_top10

```

**Findings:**

### QUESTION NO. 3: What percentage of the chemicals released in the country are transferred off-site? In which neighborhoods are the facilities located that receive these chemical transfers or disposals? Why they are released here?

#### a) Percentage transferred off-site.

```{r}
pct_offsite <- tri_clean |>
  summarise(
    total_offsite = sum(off_site_release_total, na.rm = TRUE),
    total_onsite  = sum(on_site_release_total, na.rm = TRUE),
    total_all     = sum(total_releases, na.rm = TRUE)
  ) |>
  mutate(
    percent_offsite = 100 * total_offsite / total_all,
    percent_onsite  = 100 * total_onsite / total_all
  )

print(pct_offsite)
  

```

#### b) Counties where chemicals are released.

```{r}
receiving_sites <- tri_clean |>
  filter(off_site_release_total > 0) |>
  select(facility_name, city, county, st, zip, off_site_release_total,latitude,longitude) |>
  arrange(desc(off_site_release_total))

head(receiving_sites, 10)

```

### QUESTION NO. 4: How do federally owned facilities differ from privately owned facilities in the types of chemicals they report, the industries they represent, and the forms of releases they use?

**Findings:** The results show that federally owned facilities make up a very small share of all TRI-reporting sites, and their industrial footprint is much narrower than that of private facilities. Among federal facilities, the most common sector is “Other,” representing 964 federal sites. Electric utilities follow with 90 federal facilities, while all other sectors appear only in small numbers—for example, 32 in fabricated metals, 10 in transportation equipment, and just 3 each in chemicals, hazardous waste, and printing. This limited range contrasts sharply with private facilities, where chemical manufacturing alone includes more than 17,600 reports, and major sectors such as fabricated metals, primary metals, petroleum refining, and transportation equipment each contain thousands of private facilities.

Chemical classifications differ as well. Federal facilities report many substances in the TRI “Non-Metal” and “Elemental metals” categories, and they have a comparatively high proportion of chemicals classified as PBTs (persistent, bioaccumulative toxins), with 345 federal PBT reports—almost equal to their number of non-metal TRI reports. This is notable because PBTs represent some of the most hazardous substances in the TRI. Private facilities, by contrast, report a far broader range of chemical classifications, reflecting the diversity and scale of commercial chemical use.

Release form patterns also differ clearly. Federal facilities overwhelmingly report releases in form “R” (1105 reports), with only a small number (7 reports) in form “A.” Private facilities also primarily report “R” releases, but at vastly larger scale—over 67,000 “R” reports and 8,700 “A” reports. While both groups rely on the same dominant release category, federal facilities appear to engage in fewer types of release activities overall, again reflecting their smaller and more specialized operational profile.

Overall, federal TRI activity is small, sector-concentrated, and centered around a narrow set of chemical classifications and release forms, while private industry generates nearly all TRI reports and spans the full range of industrial, chemical, and waste-management categories.

```{r}
# Separate federal vs. non-federal facilities
federal_facilities <- tri_clean |> filter(federal_facility == "YES")
private_facilities <- tri_clean |> filter(federal_facility == "NO")

# Count by industry sector
federal_industry <- federal_facilities |>
  count(industry_sector, sort = TRUE)

private_industry <- private_facilities |>
  count(industry_sector, sort = TRUE)

head(federal_industry, 10)
head(private_industry, 10)

# Chemical types reported by federal facilities
federal_chemicals <- federal_facilities |>
  count(classification, metal_category, sort = TRUE)

head(federal_chemicals, 10)

# Release form types used by federal facilities
federal_forms <- federal_facilities |>
  count(form_type, sort = TRUE)

private_forms <- private_facilities |>
  count(form_type, sort = TRUE)

federal_forms
private_forms

```

### QUESTION NO. 5: How much of the total reported releases come from chemicals classified as carcinogens, PBTs, or PFAS, and which industry sectors account for most of these high-hazard releases?

**Findings:**

The 2024 TRI data show that high-hazard chemicals—carcinogens, PBTs, and PFAS—make up a substantial share of all toxic releases reported nationwide. Out of 3.29 billion pounds of total releases, carcinogens account for 8.3 percent (274 million pounds), while PBTs account for an even larger share: 18.8 percent, or more than 620 million pounds. Together, these two categories represent more than one-quarter of all reported toxic releases.

Although PFAS releases appear extremely small by weight (0.012 percent of all TRI releases), their environmental and health risks are disproportionately high compared with their mass. TRI’s weight-based classification understates their significance because PFAS chemicals cause biological and ecological harm at tiny concentrations.

Industry sectors vary dramatically in their contribution to high-hazard releases. The chemical manufacturing sector stands out overwhelmingly, reporting more than **2.28 trillion pounds** of hazardous releases—by far the largest total of any sector. Primary metals (690 billion pounds) and metal mining (389 billion pounds) follow, reflecting the heavy-chemical and metal-processing operations central to those industries. Hazardous

### **a. Overall share of carcinogen, PBT, and PFAS releases**

```{r}
hazard_totals <- tri_clean |>
mutate(
is_carcinogen = carcinogen == "YES",
is_pbt        = pbt == "YES",
is_pfas       = pfas == "YES"
) |>
summarise(
total_releases_all   = sum(total_releases, na.rm = TRUE),
total_carcinogen     = sum(if_else(is_carcinogen, total_releases, 0), na.rm = TRUE),
total_pbt            = sum(if_else(is_pbt,        total_releases, 0), na.rm = TRUE),
total_pfas           = sum(if_else(is_pfas,       total_releases, 0), na.rm = TRUE)
) |>
mutate(
pct_carcinogen = 100 * total_carcinogen / total_releases_all,
pct_pbt        = 100 * total_pbt        / total_releases_all,
pct_pfas       = 100 * total_pfas       / total_releases_all
)

hazard_totals

```

### **b. High-hazard releases by industry sector**

```{r}
hazard_by_sector <- tri_clean |>
mutate(
is_carcinogen = carcinogen == "YES",
is_pbt        = pbt == "YES",
is_pfas       = pfas == "YES",
any_hazard    = is_carcinogen | is_pbt | is_pfas
) |>
group_by(industry_sector) |>
summarise(
total_releases        = sum(total_releases, na.rm = TRUE),
hazard_releases       = sum(if_else(any_hazard, total_releases, 0), na.rm = TRUE),
carcinogen_releases   = sum(if_else(is_carcinogen, total_releases, 0), na.rm = TRUE),
pbt_releases          = sum(if_else(is_pbt,        total_releases, 0), na.rm = TRUE),
pfas_releases         = sum(if_else(is_pfas,       total_releases, 0), na.rm = TRUE)
) |>
mutate(
pct_hazard_in_sector = 100 * hazard_releases / total_releases
) |>
arrange(desc(hazard_releases))

head(hazard_by_sector, 10)

```

### **c. Simple bar chart of sectors with most hazard releases**

```{r}
hazard_by_sector |>
slice_max(hazard_releases, n = 10) |>
ggplot(aes(x = reorder(industry_sector, hazard_releases),
y = hazard_releases)) +
geom_col() +
coord_flip() +
labs(
title = "Top 10 Industry Sectors by High-Hazard Releases (TRI 2024)",
x = "Industry sector",
y = "Releases of carcinogens, PBTs, or PFAS (lbs)"
)
```
