---
title: "Reverse Engineering Project"
author: "Quinn Muller, Dylan Schmidt and Isabella Gonzalez "
date: "Oct 22"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this notebook, we are reverse engineering the story, [Inside The Political Donation History Of Wealthy Sports Owners] (<https://fivethirtyeight.com/features/inside-the-political-donation-history-of-wealthy-sports-owners/>)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
options(scipen=999)
library(tidyverse)
library(lubridate)
library(janitor)

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data.

```{r}
sports_political_donations <- read_csv("data/sports-political-donations.csv") |> clean_names()
```

```{r}
sports_political_donations_cleaned <- sports_political_donations  |> 
  mutate(
    amount = str_replace_all(amount, "\\$|\\,", "") |> 
            as.numeric()
  )
```

## Sentences to Engineer

In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

-   **Sentence text**: American professional sports owners have contributed nearly \$47 million in federal elections since 2015, according to research by ESPN in partnership with FiveThirtyEight.
-   **Analysis summary**: We thought that the first sentence, which is the angle of the story, wouldn't be that easy to decipher, but it turns out that it is the easiest sentence to confirm. It's only the total sum of the donations.

```{r}
# American professional sports owners have contributed nearly $47 million in federal elections since 2015, according to research by ESPN in partnership with FiveThirtyEight, including $10 million to Republican causes and $1.9 million to Democratic causes so far in the 2020 election cycle.

sports_political_donations_cleaned |> 
  summarise(
    total_donations = sum(amount)
  )

donations_by_party_all  <- sports_political_donations_cleaned |> 
  filter(election_year == 2020) |> group_by(party) |> summarise(total_by_party = sum(amount, na.rm = TRUE),
    number_donations = n()
  ) 
    
```

### Sentence 2

-   **Sentence text**: That strong Republican lean is consistent with owners’ spending in the 2018 and 2016 federal elections as well. A deep search in the Federal Election Commission database of campaign finances for principal owners, controlling owners, co-owners and commissioners from the NBA, NFL, NHL, WNBA, MLB and NASCAR reveals that this deep-pocketed group has sent 34.2 million (72.9 percent) to Republican campaigns or super PACs purely supporting Republican causes, compared to 10.1 million (21.5 percent) to Democrats over the past three elections.
-   **Analysis summary**: Yes, we were able to prove the data: The Democratic party has received 10,113,639 over the past three election cycles; 21.53% of the total amount of donations over that time. The Republican party has received 34,246,016 over the past three election cycles; 72.9% of the total amount of donations over that time. The total amount of donations (denominator) includes donations to bipartisan causes, even if they have a partisan bias, and to independent causes. The partisan totals (numerator) that are used to determine the percentage represents donations to strictly Democratic or Republican entities, not to bipartisan causes that are "mostly democratic" or "mostly republican.

```{r}
#That strong Republican lean is consistent with owners’ spending in the 2018 and 2016 federal elections as well. A deep search in the Federal Election Commission database of campaign finances for principal owners, controlling owners, co-owners and commissioners from the NBA, NFL, NHL, WNBA, MLB and NASCAR reveals that this deep-pocketed group has sent 34.2 million (72.9 percent) to Republican campaigns or super PACs purely supporting Republican causes, compared to 10.1 million (21.5 percent) to Democrats over the past three elections.

league_check <- sports_political_donations_cleaned  |> 
  group_by(league) |> 
  summarize(total_amount = sum(amount))

partisan_totals <- sports_political_donations_cleaned  |> 
  group_by(party) |>
  summarize(total_amount = sum(amount))

partisan_pcts <- partisan_totals |>
  mutate(party_pct = (total_amount)/sum(total_amount) * 100)

election_check <- sports_political_donations_cleaned |>
  group_by(election_year) |>
  summarize(total_amount = sum(amount))

```

### Sentence 3

-   **Sentence text**: Drilling down deeper into the data, we can see that federal contributions in each league lean red, except for in the WNBA.
-   **Analysis summary**: Based on our findings, the writers of the article were right for the most part. All of the leagues in the dataset lean red, including the WNBA — which they listed as the only league to lean Democrat. The donations to the Republican party are significantly more than the donation total to the Democrat party, with the WNBA as the closest difference (973459 to 824640).

```{r}
#Drilling down deeper into the data, we can see that federal contributions in each league lean red, except for in the WNBA.

sports_political_donations_cleaned |> 
  filter (party == "Democrat" | party == "Republican") |> 
  group_by(league, party) |> 
  summarise(
    total_donations = sum(amount)
  )

```

### Sentence 4

-   **Sentence text**: Owners’ conservative lean extends into the current presidential cycle as well. Americans have donated similar amounts to both President Trump and former Vice President Joe Biden, but owners are giving more to Trump, according to the database.
-   **Analysis summary**: Sentence number 4 is the same that the second part of the first sentence in the story, but the reporters rephrase the results different. We used the same code in both of them because there is no column that refers to candidates only, so we assume that they are referring to Trump = Republican donations, and Biden = Democratic donations.

```{r}
#Owners’ conservative lean extends into the current presidential cycle as well. Americans have donated similar amounts to both President Trump and former Vice President Joe Biden, but owners are giving more to Trump, according to the database.

donations_by_party_all  <- sports_political_donations_cleaned |> 
  filter(election_year == 2020) |> group_by(party) |> summarise(total_by_party = sum(amount, na.rm = TRUE),
    number_donations = n()
  ) 

```

### Sentence 5

-   **Sentence text**: Which team owners gave the most? Donations from team owners, by cycle and party donated to.
-   **Analysis summary**: This sentence was difficult to find because the way the reporters group the cycles, the leagues and the total donations inside the group of the owners, team and party. It was a challenge to fin the correct path to code the list the story has at the end. That list is very attractive because we believe that from the moment you read the article's headline, you can think: "Which political party did my team's owner donate to? Or that team's owner?" So, doing the cross-reference with each one of the teams is crucial, compelling, and interesting.

```{r}
#Which team owners gave the most? Donations from team owners, by cycle and party donated to
final_list <- sports_political_donations_cleaned |> 
  group_by(owner, team, party)

print(final_list) 

donations_summary <- final_list |> 
  summarize(
    `cycle` = paste(sort(unique(na.omit(election_year))), collapse = ", "),
    `leagues` = paste(unique(na.omit(league)), collapse = ", "),
    `total donations` = sum(amount, na.rm = TRUE)
  )  |> 
  ungroup()

head(donations_summary)
```

-30-
