---
title: "Reverse Engineering Project"
author: "Quinn Muller, Dylan Schmidt and Isabella Gonzalez "
date: "Oct 22"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this notebook, we are reverse engineering the story, [Inside The Political Donation History Of Wealthy Sports Owners] (<https://fivethirtyeight.com/features/inside-the-political-donation-history-of-wealthy-sports-owners/>)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
options(scipen=999)
library(tidyverse)
library(lubridate)
library(janitor)

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data.

```{r}
sports_political_donations <- read_csv("data/sports-political-donations.csv") |> clean_names()
```

```{r}
sports_political_donations_cleaned <- sports_political_donations  |> 
  mutate(
    amount = str_replace_all(amount, "\\$|\\,", "") |> 
            as.numeric()
  )
```

## Sentences to Engineer

In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

-   **Sentence text**: American professional sports owners have contributed nearly \$47 million in federal elections since 2015, according to research by ESPN in partnership with FiveThirtyEight.
-   **Analysis summary**: We thought that the first sentence, which is the angle of the story, wouldn't be that easy to decipher, but it turns out that it is the easiest sentence to confirm. It's only the total sum of the donations.

```{r}
# American professional sports owners have contributed nearly $47 million in federal elections since 2015, according to research by ESPN in partnership with FiveThirtyEight, including $10 million to Republican causes and $1.9 million to Democratic causes so far in the 2020 election cycle.

sports_political_donations_cleaned |> 
  summarise(
    total_donations = sum(amount)
  )

donations_by_party_all  <- sports_political_donations_cleaned |> 
  filter(election_year == 2020) |> group_by(party) |> summarise(total_by_party = sum(amount, na.rm = TRUE),
    number_donations = n()
  ) 
    
```

### Sentence 2

-   **Sentence text**: That strong Republican lean is consistent with owners’ spending in the 2018 and 2016 federal elections as well. A deep search in the Federal Election Commission database of campaign finances for principal owners, controlling owners, co-owners and commissioners from the NBA, NFL, NHL, WNBA, MLB and NASCAR reveals that this deep-pocketed group has sent 34.2 million (72.9 percent) to Republican campaigns or super PACs purely supporting Republican causes, compared to 10.1 million (21.5 percent) to Democrats over the past three elections.
-   **Analysis summary**: Yes, we were able to prove the data: The Democratic party has received 10,113,639 over the past three election cycles; 21.53% of the total amount of donations over that time. The Republican party has received 34,246,016 over the past three election cycles; 72.9% of the total amount of donations over that time. The total amount of donations (denominator) includes donations to bipartisan causes, even if they have a partisan bias, and to independent causes. The partisan totals (numerator) that are used to determine the percentage represents donations to strictly Democratic or Republican entities, not to bipartisan causes that are "mostly democratic" or "mostly republican.

```{r}
#That strong Republican lean is consistent with owners’ spending in the 2018 and 2016 federal elections as well. A deep search in the Federal Election Commission database of campaign finances for principal owners, controlling owners, co-owners and commissioners from the NBA, NFL, NHL, WNBA, MLB and NASCAR reveals that this deep-pocketed group has sent 34.2 million (72.9 percent) to Republican campaigns or super PACs purely supporting Republican causes, compared to 10.1 million (21.5 percent) to Democrats over the past three elections.

league_check <- sports_political_donations_cleaned  |> 
  group_by(league) |> 
  summarize(total_amount = sum(amount))

partisan_totals <- sports_political_donations_cleaned  |> 
  group_by(party) |>
  summarize(total_amount = sum(amount))

partisan_pcts <- partisan_totals |>
  mutate(party_pct = (total_amount)/sum(total_amount) * 100)

election_check <- sports_political_donations_cleaned |>
  group_by(election_year) |>
  summarize(total_amount = sum(amount))

```

### Sentence 3

-   **Sentence text**: Drilling down deeper into the data, we can see that federal contributions in each league lean red, except for in the WNBA.
-   **Analysis summary**: Based on our findings, the writers of the article were right for the most part. All of the leagues in the dataset lean red, including the WNBA — which they listed as the only league to lean Democrat. The donations to the Republican party are significantly more than the donation total to the Democrat party, with the WNBA as the closest difference (973459 to 824640).

```{r}
#Drilling down deeper into the data, we can see that federal contributions in each league lean red, except for in the WNBA.

sports_political_donations_cleaned |> 
  filter (party == "Democrat" | party == "Republican") |> 
  group_by(league, party) |> 
  summarise(
    total_donations = sum(amount)
  )

```

### Sentence 4

-   **Sentence text**: Owners’ conservative lean extends into the current presidential cycle as well. Americans have donated similar amounts to both President Trump and former Vice President Joe Biden, but owners are giving more to Trump, according to the database.
-   **Analysis summary**: Sentence number 4 is the same that the second part of the first sentence in the story, but the reporters rephrase the results different. We used the same code in both of them because there is no column that refers to candidates only, so we assume that they are referring to Trump = Republican donations, and Biden = Democratic donations.

```{r}
#Owners’ conservative lean extends into the current presidential cycle as well. Americans have donated similar amounts to both President Trump and former Vice President Joe Biden, but owners are giving more to Trump, according to the database.

donations_by_party_all  <- sports_political_donations_cleaned |> 
  filter(election_year == 2020) |> group_by(party) |> summarise(total_by_party = sum(amount, na.rm = TRUE),
    number_donations = n()
  ) 

```

### Sentence 5

-   **Sentence text**: Which team owners gave the most? Donations from team owners, by cycle and party donated to.
-   **Analysis summary**: This sentence was difficult to find because the way the reporters group the cycles, the leagues and the total donations inside the group of the owners, team and party. It was a challenge to fin the correct path to code the list the story has at the end. That list is very attractive because we believe that from the moment you read the article's headline, you can think: "Which political party did my team's owner donate to? Or that team's owner?" So, doing the cross-reference with each one of the teams is crucial, compelling, and interesting.

```{r}
#Which team owners gave the most? Donations from team owners, by cycle and party donated to
final_list <- sports_political_donations_cleaned |> 
  group_by(owner, team, party)

print(final_list) 

donations_summary <- final_list |> 
  summarize(
    `cycle` = paste(sort(unique(na.omit(election_year))), collapse = ", "),
    `leagues` = paste(unique(na.omit(league)), collapse = ", "),
    `total donations` = sum(amount, na.rm = TRUE)
  )  |> 
  ungroup()

head(donations_summary)
```

-30-

**Key Findings and Approach**

The story's principal finding is that the majority of the donations of the owners of major U.S. sports teams favor the Republican Party.

While the data analysis itself was compelling, the reporting was notably limited in scope. The journalists primarily focused on the data, choosing not to incorporate contextual elements such as interviews with specialists, testimonies or anecdotes. Despite this, their approach proved effective because the quantitative evidence was robust; the data visualizations and statistics demonstrate the owners' inclination to donate more heavily to Republicans than Democrats. The data was used as the central tool to tell the story.

**Communication and Nut Graph**

The reporters structured the narrative by disaggregating the total donation figure, which initially served as the story's main hook. They systematically broke down the \$47 million by election year, sports league, and recipient party, effectively supporting the core argument that owners lean Republican.

However, we think that there was an opportunity for improvement in the structure of the story. Given the strength of their primary finding, the reporters could have prioritized the political leaning in the lead or nut graph of the story, stating that owners primarily donate to Republicans, rather than leading with the total amount. That figure could have then been placed in subsequent paragraphs to provide context, thus immediately focusing the reader on the most significant discovery.

The communication of specific findings was executed well. The reporters enhanced accessibility by presenting the data attractively, utilizing clear graphics and charts alongside raw numbers to simplify the complex financial distribution.

**Data and Limitations**

The data set was compiled through a search of the Federal Election Commission (FEC) database of campaign finances. The reporters had to build the data set themselves by identifying and tracking the contributions of principal owners, controlling owners, co-owners, and commissioners across the NBA, NFL, NHL, WNBA, MLB, and NASCAR. We imagine that this required significant effort in data collection and preparation prior to analysis.

The reporters transparently communicated the flaws and shortcomings of their analysis, making their work easy for readers to verify and understand. They explicitly noted that the contribution list was not "completely comprehensive" and detailed their inclusion criteria.

They also acknowledged a major limitation: the presence of a "dark figure" of potential undisclosed funding. This stems from the fact that owners, who are often millionaires and billionaires, can use tactics like having spouses or relatives make contributions to obscure their political spending. This recognition is particularly pertinent since many sports franchises are family-owned. While acknowledging this "black money," it would have been valuable if the reporters had attempted to provide a rough estimate or context for how large this unaccounted-for figure might be. Furthermore, although they claimed to "confirm" each donation's source, the methodology for this confirmation was not specified.

**Suggested Enhancements for the Story's Depth**

To avoid the story becoming tiresome due to its reliance solely on numbers, incorporating qualitative data would have been beneficial. The narrative could have been strengthened by using:

•Anecdotes and Case Studies: Including brief stories detailing a specific team owner who made a large donation to a certain candidate or party and perhaps gave them a public endorsement. This would personalize the statistics.

•Expert Opinion: Interviewing political finance specialists to provide context on why team owners donate and what influence they might seek.

•Ideological Context: Discussing whether team owners are generally known to be discreet about their political ideologies, which could contextualize their donation patterns.

We believe that the most time-consuming part of their investigation must have been the initial search for the owners and the subsequent filtration of the donation records based on their inclusion criteria. However, the data analysis itself was likely relatively straightforward. For us, it required mostly simple coding to process the created database. Therefore, the reporters likely had sufficient time to investigate and incorporate the suggested anecdotal evidence and testimonies to enhance the story's depth.

Finally, an advanced enhancement would have involved cross-referencing their donation database with public voter registration records to see if the owners had registered for a specific party's primary. This would have provided a secondary source to further substantiate the political leaning established by the donation data.

**Conclusion and Overall Assessment**

The FiveThirtyEight analysis provides a powerful, data-driven look at the political spending of elite sports team owners. Their methodology in building a custom database from FEC records and their transparency regarding its limitations demonstrate strong journalistic desire. However, we believe that relying solely on quantitative data misses an opportunity to create a richer, more engaging narrative. For future investigative work of this nature, combining the irrefutable evidence of data with the human element of storytelling would serve to both inform and captivate the audience more profoundly.
