---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse   
library(tidyverse)

# Load janitor
install.packages('janitor')
```

```{r}
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024_20250115.csv")
# Load in the Maryland grants and loan data
```

```{r}
grants <-read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022_20250115.csv")
```

```{r}
# cleaning function
cleaned_grants <- grants |>
  clean_names()

# display the cleaned dataset
cleaned_grants
```

```{r}
# cleaning function
cleaned_payments <- payments |>
  clean_names()

# display the cleaned dataset
cleaned_payments
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

```{r}
pg_schools <- cleaned_payments|>
  filter(
    str_detect(vendor_name, regex("Prince George'?s County School", ignore_case = TRUE)) |
    str_detect(vendor_name, regex("PGCPS", ignore_case = TRUE)) |
    str_detect(vendor_name, regex("P G Co Public Schools", ignore_case = TRUE)) |
    str_detect(vendor_name, regex("Prince George'?s.*Board of Ed", ignore_case = TRUE)) |
    str_detect(vendor_name, regex("PRINCE GEORGES COUNTY PUBLIC SCHOOLS", ignore_case = TRUE)) |
    str_detect(vendor_name, regex("PRINCE GEORGE'?S COMM COLLG", ignore_case = TRUE)) |
    str_detect(vendor_name, regex("PRINCE GEORGE'?S COMMUNITY COLLEGE", ignore_case = TRUE))
  )
print(pg_schools)
```

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: **147.**

```{r}
total_payments_amount <- pg_schools |>
  summarise(total_paid = sum(amount, na.rm = TRUE))
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: **2015.** **There's no data from 2016. And also, there's a huge jump between 2015 and 2016. Almost 1.2 billion dollars less.**

```{r}
payments_by_year <- pg_schools |>
  group_by(fiscal_year) |>
  summarise(
    total_paid = sum(amount, na.rm = TRUE)
  ) |>
  arrange(desc(total_paid))

cat("\n--- Payments by Fiscal Year ---\n")
print(payments_by_year)
cat("\n-------------------------------\n")
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. **I think that the cleaning we need to do is group Zip Codes when they have the same Vendor Name, for example there Medstar Family Choice with 21237 as a Zip Code and 21236. Also we would need to group names that I think have the same meaning, but are written in different ways. Examples: Baltimore City Public Schools, Baltimore Co Public Schools and Baltimore City Schools.**

```{r}
um_capital_vendors <- cleaned_payments |>
  group_by(vendor_name,vendor_zip) |>
  summarise(
    total_paid = sum(amount, na.rm = TRUE)
  ) |>
  arrange(desc(total_paid))

cat("\n--- Total Payments to Vendors for UM Capital Projects ---\n")
print(um_capital_vendors)
cat("\n--------------------------------------------------------\n")
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: **Between 2010 and 2022, stem cell research maintained an average of 7 million dollars per year. In 2023, it returned to its 2009 grant figure, which was over 10 million dollars. For me, it would be important to know how much was assigned before 2009. And also to know how much will continue to be assigned starting in 2023.**

```{r}
stem_cell_grants <- cleaned_grants |>
  filter(
    str_detect(description, regex("stem cell|stem-cell|stemcell research", ignore_case = TRUE))
  )
```

```{r}
stem_cell_funding_by_year <- stem_cell_grants |>
  group_by(fiscal_year) |>
  summarise(
    total_grants = sum(amount, na.rm = TRUE)
  ) |>
  arrange(fiscal_year)
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: **There're 17. But a lot of them are UMD just different ways to spell it. For me is interesting that UMD has a Data Center.**

```{r}
umcp_zip <- cleaned_grants |>
  filter(zip_code == "20742")
```

```{r}
umcp_zip_funding <- umcp_zip |>
  group_by(grantee,zip_code)  |>
  summarise(
    total_funding = sum(amount, na.rm = TRUE)
  )  |>
  arrange(desc(total_funding))
```
