---
title: "lab_10"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## You will need

-   Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Load libraries
library(tidyverse)
library(sf)
library(janitor)
library(tigris)

options(tigris_use_cache = TRUE)
options(tigris_class = "sf")

```

Let's explore Maryland elections, power outages and electric vehicle data and make some maps!

## Questions

**Q1.** Q1. In 2022, Maryland elected the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, calculate the total number of votes for both the governor's race and the attorney general's race, then percentages for both Moore and Brown *based on the total of their respective contests*. Then calculate the difference between their percentages.

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?

**A1.** The map shows that Moore had better results than Brown across most Maryland counties. The most interesting county is Garrett, where Brown slightly outperformed Moore despite the county being more Republican.

```{r}
md_statewide_county <- read_csv("data/md_statewide_county.csv") |> 
  clean_names() |> 
  mutate(
    geoid = as.character(geoid),
    total_ag = peroutka + brown + ag_writein,
    total_gov = cox + moore + lashar + wallace + harding + sefcik + gov_writein,
    pct_moore = (moore / total_gov) * 100,
    pct_brown = (brown / total_ag) * 100,
    moore_brown_diff = pct_moore - pct_brown
  )


md_counties_raw <- counties(cb = TRUE)

md_counties <- md_counties_raw |> 
  clean_names() |> 
  filter(statefp == "24")

md_statewide_geometry <- md_counties |> 
  inner_join(md_statewide_county, by = "geoid")

ggplot() +
  geom_sf(data = md_statewide_geometry, aes(fill = moore_brown_diff)) +
  theme_minimal()

        
```

**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland from last March. Load the `county_outages_march2023.csv` file in the data folder and calculate a rate of customers with outages per 10,000 total customers for each county. Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2.** The map shows that Talbot County has higher outage rates than the rest of Maryland. With more months of data, my next steps would be to check whether this county consistently shows elevated outages, and look patterns by date. 

```{r}
outages <- read_csv("data/county_outages_march2023.csv") |> 
  clean_names() |> 
  mutate(
    county = str_to_title(county),   # standardize names
    outage_rate_10k = (outage_customers / total_customers) * 10000
  )

md_outages_geometry <- md_counties |> 
  mutate(county = str_to_title(name)) |>    # md_counties has a 'name' column
  inner_join(outages, by = "county")

ggplot(md_outages_geometry) +
  geom_sf(aes(fill = outage_rate_10k), color = "white", size = 0.2) +
  geom_sf_text(aes(label = county), size = 3, color = "black") +
  scale_fill_viridis_b(
    option = "plasma",
    name = "Outages per 10k"
  ) +
  theme_minimal() +
  labs(
    title = "Maryland Power Outages, March 2023",
    subtitle = "Customers affected per 10,000 total customers",
    caption = "Source: county_outages_march2023.csv"
  )
```

**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change. What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3.** The second map shows that Montgomery, Howard, and Frederick counties stand out with the highest number of electric vehicles. This is different from the percentage-change map, where rural counties show the biggest changes. 

```{r}
ev <- read_csv("data/ev_2020_2023.csv") |>
  clean_names() |>
  mutate(
    county = str_to_title(county),
    pct_change = (march_2023 - july_2020) / july_2020
  )

md_ev_change <- md_counties |>
  mutate(county = str_to_title(name)) |>  
  inner_join(ev, by = "county")

ggplot(md_ev_change) +
  geom_sf(aes(fill = pct_change), color = "white", size = 0.2) +
  scale_fill_viridis_c(
    option = "mako",
    labels = scales::percent_format(accuracy = 1),
    name = "Percent change"
  ) +
  theme_minimal() +
  labs(
    title = "Percentage Change in EV & Hybrid Vehicles (2020–2023)",
    subtitle = "Percent growth from July 2020 to March 2023",
    caption = "Source: ev_2020_2023.csv"
  )

pop2021 <- get_acs(
  geography = "county",
  variables = "B01003_001",  # total population
  year = 2021,
  state = "MD",
  geometry = FALSE           # we already have geometry in md_counties
) |> 
  clean_names() |>
  mutate(
    geoid = as.character(geoid),
    population = estimate
  ) |>
  select(geoid, population)

md_ev_percap <- md_ev_change |>
  left_join(pop2021, by = "geoid") |>
  mutate(
    ev_per_1000 = (march_2023 / population) * 1000   # per 1,000 people
  )

ggplot(md_ev_percap) +
  geom_sf(aes(fill = ev_per_1000), color = "white", size = 0.2) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "EVs per 1,000 people",
    labels = scales::comma_format(accuracy = 1)
  ) +
  theme_minimal() +
  labs(
    title = "Per-Capita EV & Hybrid Vehicles (March 2023)",
    subtitle = "Vehicles per 1,000 residents",
    caption = "Source: ev_2020_2023.csv & ACS 2021"
  )
```
